Gompertzian growth as infinitely correlated non-local growth

### 2023-09-05 arxiv2/new_draft.tex
small edits

\begin{abstract}
Logistic and Gompertzian growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth. In spite of this bridge, the biological foundation of Gompertzian growth is only vaguely understood. To add to our biological understanding, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities. The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented. For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertzian growth. We thus equate Gompertzian growth with correlated microscopic particles that exhibit long-range or non-local communication. This conclusion adds to the interpretation of biological systems that undergo Gompertzian growth where long-range communication could be interpreted as the source of growth being external to the individual particles, like a field that stimulates growth simultaneously across all entities. In situations of Gompertzian decay this field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, while in situations of Gompertzian growth the field could be an internal or external electromagnetic field.
\end{abstract}

@@

Another way to generalize the logistic equation is to augment the resource terms to higher orders as follows,
\begin{equation}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta \left[(1 - X(t)) + \frac{1}{2}(1 - X(t))^2 + ... \frac{1}{K}(1 - X(t))^K\right],
\end{equation}
where fractional prefactors to each term will be justified in the sequel.
After some algebra, this augmented logistic equation reduces to a closed form expression involving Gauss' Hypergeometric function, ${}_1F_{2}$, viz.
\begin{equation}
 \label{eq:modLogistic}
\frac{d}{dt}\ln{X(t)} = - \frac{{}_{1}F_{2}({1,K+1,K+2;1-X(t)})}{K+1}(1-X(t))^{K+1} - \ln{X(t)}.
 \end{equation}

@@

As N grows, both of geometric and the arithmetic means converge in expectation to their respective distributional parameters, and thus we can identify these macroscopic quantites starting from pairwise microscopic principles. Specifically, letting $\tilde{x}(N)$ represent the sample geometric mean of $x_i$ and $\bar{x}(N)$ represent the sample arithmetic mean, each converging to their distributional counterparts $\gamma$ and $\mu$, we have 
\begin{eqnarray*}
  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=& \lim_{N\rightarrow \infty} \frac{1}{N}\sum_i^N s_i \\\
  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \tilde{x}(N) &=& \lim_{N\rightarrow \infty}(1 - \bar{x}(N)) \\
  \frac{d}{dt} \ln \gamma &=&(1 - \mu),
  % \frac{d}{dt} \ln m &=&- \ln \bar{s},
\end{eqnarray*}
where the last equality is from a convergence in expectation.

@@

\insertPngFig{order}{Exclusive interaction groups at each order: A) First order interaction, each entity is paired with a single other entity. B) Second order interaction, mutually exclusive groups of 2 entities interact with each other. B) Third order interaction, same as B) but for groups of three entities.}

Thus, let $S_k$ be the set of all mutually exclusive $k$-sized groups where $k=1,..., N/2$, where the highest order are capped at $N/2$ since that would be the largest number of entities in a single group that could interact with an identically sized group. This means that we consider the geometric mean of each $k$-group as follows,
\begin{equation}
\label{eq:modSIR4}
\frac{d}{dt} \ln x_{i_1} x_{i_2} \dots x_{i_k} = s_{j_1}s_{j_2}\dots s_{j_k}, 
\end{equation}
where the indices $i$ and $j$ represent the higher order interaction indices between the mutually exclusive groups of entities for each $k$ from 1 and up to $N/2$. Thus, the aggregated growth is represented by the sum of each interaction order,
\begin{equation}
\label{eq:MicroGomp0}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ... + \sum_{j_1, j_2, ..., j_{N/2} \in S_{N/2}} s_{j_1}s_{j_2}\dots s_{j_{N/2}},
\end{equation}
where $S_k$ is the set of all mutually exclusive $k-$sized groups. In each of the sums the total number of terms is equal to the total number of groups which at each order is $N/k$ where $k$ is the interaction order. Thus, if we divide the right hand side by $N$, we can identify a convergence property as both $k$ and $N$ grows:
\begin{enumerate}
\item As the number of entities, N, grows each term will converge in expectation to higher powers of the mean of $s_i$, multiplied by the prefactor $1$, $1/2$, $1/3$, etc., 
\begin{equation}
\label{eq:MicroGomp11}
\lim_{N\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = \frac{1}{N}\frac{N}{k} \mu^{k} = \frac{1}{k} \mu^{k}.
\end{equation}
\item As the higher order interaction terms, k, grow
\begin{equation}
\label{eq:MicroGomp12}
\lim_{k\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = 0
\end{equation}
\end{enumerate}
Thus, as each higher order term is composed of a decreasing number of summations and an increasing number of multiplications, the product of $s_i$ goes to zero since each random variable is $iid$ and bounded between 0 and 1. 
Now, using the Taylor series identity $\mu+\mu^2/2+... = -\ln(1-\mu)$, we obtain Gompertz growth in terms of the arithmetic and the geometric means of the microscopic entities
\begin{eqnarray*}
\label{eq:MicroGomp13}
\lim_{N\rightarrow \infty}\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=&  \lim_{N\rightarrow \infty}\sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ...\\
\frac{d}{dt} \ln \gamma &=&  \mu + \frac{1}{2} \mu^2 + ...\\
\frac{d}{dt} \ln \gamma &=& - \ln (1-\mu).
\end{eqnarray*}

We finally note that for the right hand side to exhibit asymptotically infinite growth at the onset of growth, it is not only sufficient but also necessary to have higher order terms. This is seen by virtue of the infinite polynomial representation of the logarithm and implies that Gompertz growth cannot result from pairwise interactions only under the assumptions presented in this model.

@@

As a side note, the Poisson distribution can be included in this discussion as the distribution which governs the infinitesimal growth rate of a coherent or infinitely correlated group of entities. By letting the growth rate be the rate at which a zero-count is observed with the Poisson mean parameter equal to $-\beta t$, one recovers the Gompertz model in Equation \ref{eq:Gomp}. This argument was put forth by \citet{shklovskii2005simple} and \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system. One the other hand, Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution. This mapping between coherence and incoherence was postulated as a medical diagnostic technique by \citet{zhang1994log}.

@@

To summarize, we have seen that both time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and infinitely correlated growth yields the Gompertz function.

We have further shown how Gompertz growth emerges both at the microscopic and the macroscopic level as a non-local effect. A novel bridge between logistic and Gompertz growth is established involving Gauss Hypergeometric function that serves as a replacement for the Richard's model derived from first principles. This bridge presupposes the source of growth to be the potential abundance as opposed to the realized abundance. On the microscopic level, the Gompertz model emerges through an exclusivity rule at every order of interaction. As a corollary, this thesis shows how the logistic function is the result of local growth processes with pairwise-interactions only, in line with its observation in the fields of predator-prey models and in communicable diseases.

@@

It is worth noting that the original formulation of the Gompertz model made by Gompertz himself was concerning mortality rates as a function of age. He noticed that mortality rates increased at a geometric rate with age. In this context the ubiquitous stressor postulated above is of course time itself. In other words, Gompertz growth arises without the need for communication across entities, but is mediated by a mutual field. One could then see the entities as communicating indirectly though this field, in a perceived one-way fashion from the field to each entity.

>>>>

\begin{abstract}
Logistic and Gompertzian growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth. In spite of this bridge, the biological foundation of Gompertzian growth is only vaguely understood. To add to our biological understanding, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities. The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented. For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertzian growth. We thus equate Gompertzian growth with correlated microscopic particles that exhibit long-range or non-local communication. This conclusion adds to the interpretation of biological systems that undergo Gompertzian growth where long-range communication could be interpreted as the source of growth being external to the individual particles, like a field that stimulates growth simultaneously across all entities. In situations of Gompertzian decay this field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, while in situations of Gompertzian growth the field could be an internal or external electromagnetic field. Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.

\end{abstract}

@@

Another way to generalize the logistic equation is to augment the resource terms to higher orders as follows,
\begin{equation}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta \left[(1 - X(t)) + \frac{1}{2}(1 - X(t))^2 + ... \frac{1}{K}(1 - X(t))^K\right],
\end{equation}
where fractional prefactors to each term will be justified in the sequel.
After some algebra and letting $\beta=1$ for simplicity, this augmented logistic equation reduces to a closed form expression involving Gauss' Hypergeometric function, ${}_1F_{2}$, viz.
\begin{equation}
 \label{eq:modLogistic}
\frac{d}{dt}\ln{X(t)} = - \frac{{}_{1}F_{2}({1,K+1,K+2;1-X(t)})}{K+1}(1-X(t))^{K+1} - \ln{X(t)}.
 \end{equation}

@@

As N grows, both of geometric and the arithmetic means converge in expectation to their respective distributional parameters, and thus we can identify these macroscopic quantites starting from pairwise microscopic principles. Specifically, letting $\tilde{x}(N)$ represent the sample geometric mean of $x_i$ and $\bar{x}(N)$ represent the sample arithmetic mean, each converging to their distributional counterparts $\gamma$ and $\mu$, we have 
\begin{eqnarray*}
  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=& \lim_{N\rightarrow \infty} \frac{1}{N}\sum_i^N s_i \\\
  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \tilde{x}(N) &=& \lim_{N\rightarrow \infty}(1 - \bar{x}(N)) \\
  \frac{d}{dt} \ln \gamma &=&(1 - \mu),
  % \frac{d}{dt} \ln m &=&- \ln \bar{s},
\end{eqnarray*}
where the last equality follows from a convergence in expectation.

@@

\insertPngFig{order}{Exclusive interaction groups at each order: A) First order interaction, each entity is paired with a single other entity. B) Second order interaction, mutually exclusive groups of two entities interact with each other. B) Third order interaction, same as B) but for groups of three entities.}

Thus, let $S_k$ be the set of all mutually exclusive $k$-sized groups where $k=1,..., N/2$, where the highest order are capped at $N/2$ since that would be the largest number of entities in a single group that could interact with an identically sized group. This means that we consider the geometric mean of each $k$-group as follows,
\begin{equation}
\label{eq:modSIR4}
\frac{d}{dt} \ln x_{i_1} x_{i_2} \dots x_{i_k} = s_{j_1}s_{j_2}\dots s_{j_k}, 
\end{equation}
where the indices $i$ and $j$ represent the higher order interaction indices between the mutually exclusive groups of entities for each $k$ from 1 and up to $N/2$. Thus, the aggregated growth is represented by the sum of each interaction order,
\begin{equation}
\label{eq:MicroGomp0}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ... + \sum_{j_1, j_2, ..., j_{N/2} \in S_{N/2}} s_{j_1}s_{j_2}\dots s_{j_{N/2}},
\end{equation}
where $S_k$ is the set of all mutually exclusive $k-$sized groups. In each of the sums the total number of terms is equal to the total number of groups which at each order is $N/k$ where $k$ is the interaction order. Thus, if we divide the right hand side by $N$, we can identify a convergence property as both $k$ and $N$ grows:
\begin{enumerate}
\item As the number of entities grow each term will converge in expectation to higher powers of the mean of $s_i$, multiplied by the prefactor $1$, $1/2$, $1/3$, etc., 
\begin{equation}
\label{eq:MicroGomp11}
\lim_{N\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = \frac{1}{N}\frac{N}{k} \mu^{k} = \frac{1}{k} \mu^{k}.
\end{equation}
\item As the higher order interaction terms grow
\begin{equation}
\label{eq:MicroGomp12}
\lim_{k\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = 0
\end{equation}
\end{enumerate}
Thus, as each higher order term is composed of a decreasing number of summations and an increasing number of multiplications, the product of $s_i$ goes to zero since each random variable is $iid$ and bounded between 0 and 1. 
Now, using the Taylor series identity $\mu+\mu^2/2+... = -\ln(1-\mu)$, we obtain Gompertz growth in terms of the arithmetic and the geometric means of the microscopic entities
\begin{eqnarray*}
\label{eq:MicroGomp13}
\lim_{N\rightarrow \infty}\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=&  \lim_{N\rightarrow \infty}\left[\sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ...\right]\\
\frac{d}{dt} \ln \gamma &=&  \mu + \frac{1}{2} \mu^2 + ...\\
\frac{d}{dt} \ln \gamma &=& - \ln (1-\mu).
\end{eqnarray*}

We finally note that for the right hand side to exhibit asymptotically infinite growth at the onset of growth, it is not only sufficient but also necessary to have higher order terms. This is seen by virtue of the infinite polynomial representation of the logarithm and implies that Gompertz growth cannot result from pairwise interactions under the assumptions presented in this model, but rather requires all higher order interactions to be present.

@@

In general, our choice of time as the dependent variable can be substituted for any suitable metric of choice. Another example of space as the dependent variable in the Gompertz setting is given by \citet{molski2003coherent}.

As a side note, the Poisson distribution can be included in this discussion as the distribution which governs the infinitesimal growth rate of a coherent or infinitely correlated group of entities. By letting the growth rate be the rate at which a zero-count is observed with the Poisson mean parameter equal to $-\beta t$, one recovers the Gompertz model in Equation \ref{eq:Gomp}. This argument was put forth by \citet{shklovskii2005simple}, and \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system. One the other hand, Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution which is in accordance with our results and discussion here. The mapping between coherence and incoherence has also been postulated as a medical diagnostic technique by \citet{zhang1994log}, where incoherence is a sign of disease while coherence is a sign of well-being.

@@

To summarize, we have seen that both time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.

We have further shown how Gompertz growth emerges both at the microscopic and the macroscopic level as a non-local effect. A novel bridge between logistic and Gompertz growth is established involving Gauss Hypergeometric function that serves as a replacement for the Richard's model and is derived from first principles. This bridge presupposes the source of growth to be the potential abundance as opposed to the realized abundance. On the microscopic level, the Gompertz model emerges through an exclusivity rule at every order of interaction. As a corollary, this thesis shows how the logistic function is the result of local growth processes with pairwise-interactions only, in line with its observation in the fields of predator-prey models and in communicable diseases.

@@

It is worth noting that the original formulation of the Gompertz model made by Gompertz himself was concerning mortality rates as a function of age. He noticed that mortality rates increased at a geometric rate with age. In this context the ubiquitous stressor postulated above is of course time itself and thus Gompertz growth in this context will be implicitly \textit{time dependent}. In other words, Gompertz growth arises without the need for communication across entities, but is mediated by a mutual field. One could then see the entities as communicating indirectly though this field, in a perceived one-way fashion from the field to each entity.
